Utilizing Parallel Workers: LLVM's Vectorization Plan

### 2025-01-28 paper/paper.tex
paper

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}

@@

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop must be executed due to 
unclear data dependencies.

@@

Intel's initial VPlan proposal aimed to unite any 
auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a 
single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
This system would use VPlans to represent different 
vectorization approaches, abstracting away from the underlying scalar loop IR using its own 
hierarchical control flow graph (H-CFG). This allows iteratively exploring different 
optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
Each VPlan could to calculate its cost and execute itself, materializing the vectorization 
approach back into LLVM IR.

@@

\subsection{VPlan Structure}
\emergencystretch 2em
Before explaining how far this system has been implemented in LLVM, we will review
the different elements of a \texttt{VPlan} in more detail.

@@

\paragraph{H-CFG}
This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main elements of a single VPlan. It is hierarchical
in that a single node of this graph can be a complete subgraph, representing more complex nested
behavior.
Specifically, any node of this H-CFG is one of three types:

@@

Inside \texttt{vector loop} is only a single \texttt{VPBasicBlock} node \texttt{vector.body}. This block 
models the entire vectorized loop via \texttt{VPRecipes}.

\paragraph{Recipes}

@@

\section{State of VPlan (LLVM 19.1.4)}
\emergencystretch 3em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen some development, any 
outer loop vectorization features are still experimental and must be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

@@

One can find more details about these legality checks and their implementations by looking 
at the LLVM codebase under \texttt{LoopVectorizationLegality::\allowbreak canVectorize}.

@@

\paragraph{Cost Model}
\emergencystretch 4em
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally lookup tables influenced by the chosen target 
architecture.

The before-mentioned \texttt{LoopVectorizationCostModel} is a remainder of the old \texttt{LoopVectorize} 
implementation before VPlan. It is a single object that computes different costs based on the scalar loop 
IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
after vectorization or the LLVM values for which the cost can 
be ignored. The latter includes, for instance, loop invariant operations that would be hoisted out of the loop 
or trivially dead instructions.

@@

A \texttt{VPTransformState} object is passed along to store information 
needed for generating the output IR, such as the chosen vectorization factor or generated values that were 
used by later \texttt{VPRecipes}.

\begin{figure}
  \centering
  \includegraphics[width=0.35\textwidth]{images/inner-loop-vec-loop-cfg-color.png}
  \caption{The generalized CFG for a vectorized loop with trip count \texttt{N}, 
  iteration variable \texttt{i}, vectorization factor \texttt{VF}, 
  and unroll factor \texttt{UF} in LLVM. A VPlan would so far explicitly model the red-highlighted 
  region in its H-CFG. If a node branches, the condition is stated in the body, and the two possible 
  path edges are marked with T (True) and F (False).}
  \label{fig:inner-loop-vec-cfg}
\end{figure}

@@

Additionally, before the initial VPlans are constructed at the beginning of Phase 2, it is also checked if 
the scalar epilogue loop could be folded into the vectorized loop version itself by increasing the vector 
trip count over the actual loop trip count and masking all operations. This tail-folding approach in LLVM 
has not yet been enabled by default.

@@

\subsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
abstracted scalar IR instructions. Besides that, the community also focuses on enabling tail folding by
default, which would lead to a significant performance increase in some instances.

@@

\subsection{Function Vectorization}
There is currently no active development for whole-function vectorization in the
llvm-project's upstream.

@@

\section{Related Work}
\label{sec:relatedwork}
Next to the main llvm-projects upstream, many projects are using or extending the LLVM toolchain.

@@

To achieve this, RV implements partial control flow linearization~\cite{10.1145/3192366.3192413}
and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
implement this control flow linearization and divergence analysis into LLVM's 
upstream~\cite{rvproposal,rvproposaldep}.

% TODO: Dependency Analysis (SCEV Expressions)
%         - https://www.npopov.com/2023/10/03/LLVM-Scalar-evolution.html
%         - https://www.youtube.com/watch?v=AmjliNp0_00
% TODO: Polly and its use for loop transformations affecting data dependencies
% TODO: Comparisons LLVM with GCC
%         - Tail loop folding (https://github.com/llvm/llvm-project/issues/123069)
%         - Another benchmark: https://www.spec.org/cpu2017/Docs/benchmarks/525.x264_r.html

@@

\section{Summary and Future Work}
\label{sec:summary}
Compilers have been using auto-vectorization to help developers optimize their code 
across platforms by making the best possible use of a target's SIMD capabilities. Over the years, 
LLVM's original auto-vectorization system evolved into an inflexible pipeline, making it hard 
to adapt to new features like outer-loop vectorization.

@@

Outer loop vectorization support via VPlan is still experimental and separated from the inner loop
LV integration. In the near future, development on outer loop vectorization will continue, and merging 
it with LV's VPlan path is planned.

Active development of other approaches, such as SLP and whole-function vectorization based on the VPlan 
architecture, has yet to be started.

% TODO: Write about potential future work benchmarking
%       Loop Code Benchmark collection
%       Quick note about collaborative development (example stability & miscompile policy)
%       LLVM Loop Optimization as one of the most active subareas of LLVM development

>>>>

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage{flushend}
\usepackage{xurl}
\usepackage[babel]{csquotes}

@@

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
epilogue/tail of the loop might be used to handle the remaining loop iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop must be executed due to 
unclear data dependencies.

@@

Intel's initial VPlan proposal aimed to unite any 
auto-vectorization systems (SLP, Inner/Outer-Loop, Whole-Function) into a 
single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
This system would use VPlans to represent different 
vectorization approaches, abstracting away from the underlying scalar loop IR using its own 
hierarchical control flow graph (H-CFG). This allows iteratively exploring different 
optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
Each VPlan could to calculate its cost and execute itself, materializing the vectorization 
approach back into LLVM IR.

@@

\subsection{VPlan Structure}
Before explaining how far this system has been implemented in LLVM, we will review
the different elements of a \texttt{VPlan} in more detail.

@@

\paragraph{H-CFG}
\emergencystretch 1em

This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main elements of a single VPlan. It is hierarchical
in that a single node of this graph can be a complete subgraph, representing more complex nested
behavior.
Specifically, any node of this H-CFG is one of three types:

@@

Inside the \texttt{vector loop} region is only a single \texttt{VPBasicBlock} node \texttt{vector.body}. This block 
models the entire vectorized loop via \texttt{VPRecipes}.

\paragraph{Recipes}
\emergencystretch 2em

@@

\section{State of VPlan (LLVM 19.1.4)}
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen some development, any 
outer loop vectorization features are still experimental and must be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

@@

@@

\paragraph{Cost Model}
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally lookup tables influenced by the chosen target 
architecture.

A remainder of the old \texttt{LoopVectorize} implementation before VPlan is the 
before-mentioned \texttt{LoopVectorizationCostModel}. It is a single object that computes different costs based on the scalar loop 
IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
after vectorization or the LLVM values for which the cost can 
be ignored. The latter includes, for instance, loop invariant operations that would be hoisted out of the loop 
or trivially dead instructions.

@@

A \texttt{VPTransformState} object is passed along to store information 
needed for generating the output IR, such as the chosen Vectorization Factor or generated values that were 
used by later \texttt{VPRecipes}.

\begin{figure}
  \centering
  \includegraphics[width=0.35\textwidth]{images/inner-loop-vec-loop-cfg-color.png}
  \caption{The generalized CFG for a vectorized loop with trip count \texttt{N}, 
  iteration variable \texttt{i}, Vectorization Factor \texttt{VF}, 
  and Unroll Factor \texttt{UF} in LLVM. A VPlan would so far explicitly model the red-highlighted 
  region in its H-CFG. If a node branches, the condition is stated in the body, and the two possible 
  path edges are marked with T (True) and F (False).}
  \label{fig:inner-loop-vec-cfg}
\end{figure}

@@

Additionally, before the initial VPlans are constructed at the beginning of Phase 2, it is also checked if 
the scalar epilogue loop could be folded into the vectorized loop version itself by increasing the vector 
trip count over the actual loop trip count and masking all operations. This tail-loop folding approach in LLVM 
has not yet been enabled by default.

@@

\subsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
abstracted scalar IR instructions. Besides that, the community also focuses on enabling tail-loop folding by
default, which would lead to a significant performance increase in some instances.

@@

\subsection{Function Vectorization}
There seems to be currently no active development on whole-function vectorization in the
llvm-project's upstream.

@@

\section{Related Work}
\label{sec:relatedwork}
\subsection{Region Vectorizers}
Next to the main llvm-projects upstream, many projects are using or extending the LLVM toolchain.

@@

To achieve this, RV implements partial control flow linearization~\cite{Moll_2021,10.1145/3192366.3192413}
and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
implement this control flow linearization and divergence analysis into LLVM's 
upstream~\cite{rvproposal,rvproposaldep}.

In a later paper, Moll et al. further generalized this region vectorizer to be able to vectorize 
nested loops in a hybrid/multi-dimensional way, pushing optimization capabilities even 
further~\cite{10.1145/3303117.3306172}.

\subsection{Comparison: GCC and LLVM}
As another popular compiler, GCC has implemented and consistently improved its auto-vectorization 
over the years~\cite{autovecoptgcc,gccvecsum}. Compared to LLVM, GCC also supports 
vectorization of outer loops and tail-loop folding by default.

In recent years, different research has been done about how compilers, especially GCC and 
LLVM, compare in terms of auto-vectorization.

Feng et al. compared multiple versions of GCC, LLVM, and Intel's ICC compiler in auto-vectorizing the 
TSVC benchmark set targeting x86-64~\cite{feng2021evaluation}. They found that in terms of compilation time, GCC 9
has much less overhead ($\approx4\%$) when auto-vectorization is enabled compared to 
LLVM 9 ($\approx24\%$). In both cases, LLVM 9 takes twice as long to compile than GCC 9. 
ICC 19 sits right between GCC 9 and LLVM 9 regarding compilation time.

While in terms of runtime, GCC 9 and LLVM 9 take around 
the same time with auto-vectorization disabled, enabling it improves GCC 9's performance by around $44\%$, 
while for LLVM 9 it is only $12\%$. ICC 19 has a better runtime than GCC 9 and LLVM 9 in both the 
non-vectorized and the auto-vectorized case.

The authors also investigated the number of successfully auto-vectorized code routines inside the TSVC 
benchmark divided into different vectorization categories. Out of a total of 151 different routines, GCC 9 
vectorized 68, LLVM 9 vectorized 49, and ICC 19 did 98.

Feng et al. concluded that auto-vectorizers were mostly limited by analysis capabilities 
needed to reason about specific optimization approaches and inaccurate cost models.

While LLVM 9 performs poorly in this paper, it is important to notice that at the time of LLVM 9, VPlan was relatively 
early into development ($\approx3$ years) and was still missing 
support for a lot of commonly used vectorization idioms such as reductions, for example.

In a more recent auto-vectorization evaluation comparing GCC 14 with LLVM 17 on the similar TSVC 2 
benchmark targeting RISC-V, the author investigated the produced vectorized code 
and drew comparisons to similar papers~\cite{gccllvmveccomp}. The SVE and RISC-V 
vector extensions are  
vector-length agnostic, 
meaning that, unlike SSE or AVX, vector instructions are not tied to specific vector register sizes. 
This makes the code more compatible and scalable across different underlying hardware.

In the paper, the author concludes that the achieved performance uplifts by auto-vectorization are very 
similar between GCC 14 and LLVM 17, with GCC only being marginally better. LLVM 17 tends to favor 
generating vector code targeting specific vector sizes. However, it is unclear whether this is an 
actual drawback.

Finally, as mentioned before, the LLVM auto-vectorization community is currently working on 
enabling tail-loop folding by default after noticing a $\approx24\%$ performance difference between the latest 
versions of GCC and LLVM in a SPEC CPU 2017 benchmark targeting RISC-V~\cite{riscvtailfolding}.

@@

\section{Summary and Future Work}
\label{sec:summary}
Compilers have been using auto-vectorization to help developers optimize their code 
across platforms by making the best possible use of a target's SIMD capabilities. Over the years, 
LLVM's original auto-vectorization system evolved into an inflexible pipeline, making it hard 
to adapt to new features like outer loop vectorization.

@@

Outer loop vectorization support via VPlan is still experimental and separated from the inner loop
LV integration. In the near future, development on outer loop vectorization will continue, and merging 
it with LV's VPlan path is planned. Otherwise, development is currently underway to enable 
tail-loop folding by default.

Active development of other approaches, such as SLP and whole-function vectorization based on the VPlan 
architecture, has yet to be started. Lastly, the initially planned cost model which enabled a standalone 
VPlan to calculate its own cost has not been implemented yet but could become a vital system in the future 
of VPlan.

In the future, it might be interesting to benchmark and compare different auto-vectorization systems, 
such as the old LLVM LoopVectorize pass, the current state of LLVM with its new VPlan system, 
GCC~\cite{autovecoptgcc}, ICC, and region vectorizers~\cite{Moll_2021,10.1145/3303117.3306172,rv}. Interesting metrics to 
compare would be compile time, performance and size of the produced code, differences in the chosen 
vectorization approaches, differences in legality checks that lead to loops not being vectorized, 
correctness, and stability. These metrics can then differ based on the target architecture.

Various suitable benchmark sets exist that have also been used for similar 
comparisons, such as TSVC 2 or SPEC CPU 2017.
